Public Health and Economic Impact of Weather events in the US
=============================================================

## Synopsis
 **Here** 

## Data Processing

1. **Note 1:** Dependencies: no
2. **Note 2:** The source documentation for this analysis is given in [NWSI][1]

For out analysis, we are going to use the NOAA Storm Database. So first we need to download it to a temporal file, expand it and put it in a data frame called weather_dataset:

```{r cache=TRUE}
filename = "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
tempfile <- tempfile()
download.file(filename, tempfile)
weather_dataset = read.csv(bzfile(tempfile), sep=",", header=T)
unlink(tempfile)
```

With the following function lets check if there are any NA's in the dataset:


```{r}
nacols <- function(df) {
    colnames(df)[unlist(lapply(df, function(x) any(is.na(x))))]
}
na_cols = nacols(weather_dataset)
na_cols
```

As we can see there are 4 columns that contain NA values.
So lets keep in mind this just in case we have to use them.

We will also check the number of rows with NA's, to have an idea of the completeness of our dataset:


```{r}
ok = complete.cases(weather_dataset)
na_rows = sum(!ok)
na_rows
```

As we can see, there are a lot (902297) of missing values in this dataset.

Lets get to know a bit out dataset. These are the fields:

```{r}
str(weather_dataset)
```

This are the field names:

```{r}
colnames(weather_dataset)
```

and this is a simple summary:

```{r}
summary(weather_dataset)
```

We have to provide a unique standard unit for the values of crop damage and property damage. I choose K (thousand's of $)

```{r}
# provide a unique standard unit for prop damage
standarizePropDmgUnit = function(propDmg, propDmgExp) {  
  if(propDmgExp=="b") {#billion
    propDmg * 1000 * 1000
    
  } else if(propDmgExp=="M") {#million
    propDmg * 1000
  } else if(propDmgExp=="m") {#Thousandth
    propDmg / (1000^6)
  } else if(propDmgExp=="H") {#hundred
    propDmg / 1000
  } else {# fr K and all other values, return as is
    #Note: its very obscure the symbol h and the numbers,
    # -, + and ?
    # I just keep it as is
    propDmg
  }
}

# provides a unique standard unit for
# crop damage values
standarizeCropDmgUnit = function(cropDmg, cropDmgExp) {  
  if(cropDmgExp=="b") {#billion
    cropDmg * 1000 * 1000
  } else if(cropDmgExp=="M") {#million
    cropDmg * 1000
  } else if(cropDmgExp=="m") {#Thousandth
    cropDmg / (1000^6)
  } else if(cropDmgExp=="H") {#hundred
    cropDmg / 1000 
  } else {# fr K and all other values, return as is
    #Note: its very obscure the symbol h and the numbers,
    # -, + and ?
    # I just keep it as is
    cropDmg
  }
}

#remove entries with 0 fatalities (harmless)
weather_dataset = weather_dataset[weather_dataset$FATALITIES != 0,]

#remove entries with no injuries (harmless)
weather_dataset = weather_dataset[weather_dataset$INJURIES != 0,]

#remove entries with no prop damage expenditures
weather_dataset = weather_dataset[weather_dataset$PROPDMG != 0,]

#remove entries with no crop damage expenditures
weather_dataset = weather_dataset[weather_dataset$CROPDMG != 0,]

for(i in 1:nrow(weather_dataset)) {
    propDmg = weather_dataset[i,"PROPDMG"]
    propDmgExp = weather_dataset[i,"PROPDMGEXP"]
    cropDmg = weather_dataset[i,"CROPDMG"]
    cropDmgExp = weather_dataset[i,"CROPDMGEXP"]
    
    weather_dataset[i,"PROPDMG"] = standarizePropDmgUnit(
      propDmg, propDmgExp)
    weather_dataset[i,"CROPDMG"] = standarizeCropDmgUnit(
      cropDmg, cropDmgExp)
    # do stuff with row
}


# IMPORTANT: please not I didn't "clean" the fields in the sense
# that I didn't merge fields together like others did.
# I tihink that for doing that, one should have more info
# on why those fields that look the same should be merged
# with confidence. else one may be twisting results

```



## Results 

We want to answer the following 2 fundamental questions:

1. Across the United States, which types of events are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

### Most harmful events for population health

The field for the event types is EVTYPE. 
Lets take a look at some event types:

```{r}
str(weather_dataset$EVTYPE)
```

In section 7 of the NWSI reference document ([NWSI][1]) we can inspect the different types of events in detail.
For example From that document we can see that Excessive Heat (Z) event type is used for reporting fatalities (directly-related) or major impacts to human health occurring during excessive heat;
Here is the quote:
**Excessive Heat (Z)**
**Fatalities (directly-related) or major impacts to human health occurring during excessive heat warning conditions are reported using this event category.**

coming back to our question, We need a meassure of the impact of each of these event types to public health. From the doc and the dataset, that impact should be given by two fields: FATALITIES (death cases) and INJURIES.

We should treat these 2 in a separate way as they are not the same. But we can "join" them together to see the overall "health impact" as this is what we want to answer

Total deaths per event type (first 10 greatest by # of deaths):
```{r}
deathsPerEvtyp = aggregate(weather_dataset['FATALITIES'],
  by=list(event = weather_dataset$EVTYPE), FUN=sum)
# order the results
deathsPerEvtyp = deathsPerEvtyp[with(deathsPerEvtyp, order(-FATALITIES)), ]

top10Fatalities = head(deathsPerEvtyp, n=10)
top10Fatalities
```

Those are the top-10 most deathful events. As we can see, tornados and excessive heat are the most fatal events by far, with 5633 and 1903 number of deaths respectively.

Total injuries per event type (first 10 greatest by # of injuries):
```{r}
injPerEvtyp = aggregate(weather_dataset['INJURIES'], 
  by=list(event = weather_dataset$EVTYPE), FUN=sum)
# order the results
injPerEvtyp = injPerEvtyp[with(injPerEvtyp, order(-INJURIES)), ]

top10Injuries = head(injPerEvtyp, n=10)
top10Injuries
```

Those are the top-10 most harmful (only injuries) events, with tornados been the most harmful (91346 injurie cases) events by far.

The following is a graph (bar plot) of the top-10 most harmful events in each case (death and injuries):

```{r}
par(mfrow = c(2, 1))
# deaths plot
barplot(top10Fatalities$FATALITIES,main="Top-10 Deaths per event type", xlab="Event type", ylab="# deaths",col=c("red", "brown","brown1","brown2", "brown3","burlywood","burlywood1","burlywood2", "burlywood3","burlywood4"), names.arg=1:10, legend=top10Fatalities[,"event"], args.legend=c(cex=0.5))
#args.legend=c(cex=0.4))
# and this is the injuries plot
barplot(top10Injuries$INJURIES,main="Top-10 Injuries count per event type", 
    xlab="Event type", ylab="# injuries",col=c("red", "brown","brown1","brown2", "brown3","burlywood","burlywood1","burlywood2", "burlywood3","burlywood4"), names.arg=1:10, legend=top10Injuries[,"event"], args.legend=c(cex=0.5))
```

And lets see the overall harm (deaths + injuries)
```{r}
# first lets merge the 2 datasets
harmPerEvtyp = merge(top10Fatalities,top10Injuries)
harmPerEvtyp$HARM = harmPerEvtyp$FATALITIES + harmPerEvtyp$INJURIES 
# order the results
harmPerEvtyp = harmPerEvtyp[with(harmPerEvtyp, order(-HARM)), ]

barplot(harmPerEvtyp$HARM,main="Top-10 harmful events", 
    xlab="Event type",ylab="harm count", col=c("red", "brown","brown1","brown2", "brown3","burlywood","burlywood1"), legend=harmPerEvtyp[,"event"],
    args.legend=c(cex=0.8))

```

As we can see, overall, the 10-most harmful event is the tornado, following is excessive heat, wind, flood, lightening, heat and flash flood. 

**We should take very special care regarding tornados!!!**

### Events with greatest economic consequences

Now lets see what happens with the economic  aspect.
The question is:
**Across the United States, which types of events have the greatest economic consequences?**
Let's see for each economic factor in each own:

The fields we are interested in are:

1."PROPDMG" (property damage)  
2."CROPDMG" (crop damage)

with corresponding units:
"PROPDMGEXP" (unit for property damage)
"CROPDMGEXP" (unit for crop damage)

So this is the property damage per event type:
```{r}
propDmgPerEvtyp = aggregate(weather_dataset['PROPDMG'],
  by=list(event = weather_dataset$EVTYPE), FUN=sum)

# order the results
propDmgPerEvtyp = propDmgPerEvtyp[with(propDmgPerEvtyp, order(-PROPDMG)), ]
top10propDmg = head(propDmgPerEvtyp, n=10)
top10propDmg
```

As we can see, the tornado and flash wind are the events whith the greatest damage

Lets see what about the crop damage
```{r}
cropDmgPerEvtyp = aggregate(weather_dataset['CROPDMG'],
  by=list(event = weather_dataset$EVTYPE), FUN=sum)

# order the results
cropDmgPerEvtyp = cropDmgPerEvtyp[with(cropDmgPerEvtyp, order(-CROPDMG)), ]

top10cropDmg = head(cropDmgPerEvtyp, n=10)
top10cropDmg
```

And we can see that Hail is the event with the greatest damage, following is flsh flood, wind, tornado, etc.

Let see a plot of all this:

```{r}
par(mfrow = c(2, 1))
# prop plot
barplot(top10propDmg$PROPDMG,main="Top-10 prop damage per event type", 
    xlab="Event type", ylab="prop damage",col=c("red", "brown","brown1","brown2", "brown3","burlywood","burlywood1","burlywood2", "burlywood3","burlywood4"), names.arg=1:10, legend=top10propDmg[,"event"],args.legend=c(cex=0.5))
# and this is the crop plot
barplot(top10cropDmg$CROPDMG,main="Top-10 Crop damage per event type", 
    xlab="Event type", ylab="crop damage",col=c("red", "brown","brown1","brown2", "brown3","burlywood","burlywood1","burlywood2", "burlywood3","burlywood4"), names.arg=1:10, legend=top10cropDmg[,"event"],args.legend=c(cex=0.5))
```


[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf        "NWSI"
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2   "Dataset"
